---
title: "Linear Models with R - Exercises"
author: "Rachel Greenlee"
date: "8/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('faraway')
```

## Exercise 1.1
The dataset teengamb concerns a study of teenage gambling in Britain. Make a
numerical and graphical summary of the data, commenting on any features that
you find interesting. Limit the output you present to a quantity that a busy reader
would find sufficient to get a basic understanding of the data.


```{r}
head(teengamb)
summary(teengamb)
```



## Exercise 1.3
The dataset prostate is from a study on 97 men with prostate cancer who were
due to receive a radical prostatectomy. Make a numerical and graphical summary
of the data as in the first question.



## Exercise 1.4
The dataset sat comes from a study entitled “Getting What You Pay For: The Debate Over Equity in Public School Expenditures.” Make a numerical and graphical
summary of the data as in the first question




## Exercise 1.5
The dataset divusa contains data on divorces in the United States from 1920 to
1996. Make a numerical and graphical summary of the data as in the first question.